Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-11-06 phyloglmm_ms.tex
temp pushing, glmmTMB fits are missing

Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
In response, researchers are studying more complicated systems and fitting more models to larger data sets describing species occurrence and traits.
One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting}. 
Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; researchers may be interested in controlling for the statistical effects of phylogenetic correlation, quantifying the evolutionary signals that underly traits or both.
Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary} 
%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
In such cases, researchers may try to find ways to simplify their analyses (i.e. controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
% \bmb{nice!}

@@

\mli{move example to the supp}
% For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
% 
% \begin{align}
% \bX = \begin{bmatrix}
% 1 & t_1  \\ 
% 1 & t_2  \\ 
% 1 & t_3 
% \end{bmatrix} 
% \end{align}
% then the term-wise phylogenetic random effects model matrix is,
% 
% \begin{align}
% \bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
% \Bigg[
% \Bigg(
% \begin{bmatrix}
% \ell_{adj1} & \ell_{adj1}  & 0 \\
% \ell_{adj2} &  0  & 0 \\
% 0  &  \ell_{adj3} & 0 \\
% 0 & 0 &  \ell_{adj4} 
% \end{bmatrix}
% \begin{bmatrix}
% 1 & 0  & 0 \\
% 0 & 1  & 0 \\
% 0 & 0  & 1  
% \end{bmatrix}
% \Bigg)
% \ast
% \begin{bmatrix}
% 1   & 1   & 1  \\ 
% t_1 & t_2 & t_3
% \end{bmatrix} 
% \Bigg]^T
% \\
% = \begin{bmatrix}
% \ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
% \ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
% \end{bmatrix}
% \end{align}

@@

\subsection{Simulation and evaluations}
We simulated 100 phylogenetic trees for each sample size (25,50,100) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
All model variants via different platforms were used to fit each realization.
Tables \ref{model_variant} shows the parameters that are estimable for each model/platform. 
We combined two convergence criteria to assess convergence for the main parameters (R0, Peff, Prep): we required a value of the
Gelman and Rubin statistic and an effective sample size (ESS) greater than 400 for each replication. For
each replication we sample four chains starting with 4000 iterations; w

@@

The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$). 
Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.

In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation. 
$\beta_0$ estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$. 
\mli{double check this, what happens when X is also phylogenetically correlated?}

@@

Multiple site fits on the other hand are a lot more similar across platforms (as there are less deviates between the fitted and simulation model) compared to the single site fits.
Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
\mli{either keep it or take it out, because  will affect other parameters}
The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.  
\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
\mli{save it for random slope discussion}

However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
Lastly, it takes $~$ 2000 seconds for lme4 to fit 1000 species model and it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
\mli{put a lower bound to give up}

@@

We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to \pkg{pez}. 
Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.

>>>>

Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
In response, researchers are studying more complicated systems and fitting more models to larger data sets describing species occurrence and traits.
One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting}. 
Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; researchers may be interested in controlling for the statistical effects of phylogenetic correlation, quantifying the evolutionary signals that underly traits or both.
Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications such as biodiversity conservation, effects of climate change, and many more \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}. 
%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
In such cases, researchers may try to find ways to simplify their analyses (i.e. controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
% \bmb{nice!}

@@

\mli{move example to the supp}
For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,

\begin{align}
\bX = \begin{bmatrix}
1 & t_1  \\ 
1 & t_2  \\ 
1 & t_3 
\end{bmatrix} 
\end{align}
then the term-wise phylogenetic random effects model matrix is,

\begin{align}
\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
\Bigg[
\Bigg(
\begin{bmatrix}
\ell_{adj1} & \ell_{adj1}  & 0 \\
\ell_{adj2} &  0  & 0 \\
0  &  \ell_{adj3} & 0 \\
0 & 0 &  \ell_{adj4} 
\end{bmatrix}
\begin{bmatrix}
1 & 0  & 0 \\
0 & 1  & 0 \\
0 & 0  & 1  
\end{bmatrix}
\Bigg)
\ast
\begin{bmatrix}
1   & 1   & 1  \\ 
t_1 & t_2 & t_3
\end{bmatrix} 
\Bigg]^T
\\
= \begin{bmatrix}
\ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
\ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
\end{bmatrix}
\end{align}

@@

\subsection{Simulation and evaluations}
We simulated 100 phylogenetic trees for each sample size (25,50,100, 500) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
All model variants via different platforms were used to fit each realization.
Tables \ref{model_variant} shows the parameters that are estimable for each platform. 
All model must pass its respective convergence test before evaluating goodness of fit. 
For bayesian GLMM, we combined two convergence criteria to assess convergence: we required a value of the
Gelman and Rubin statistic and an effective sample size (ESS) greater than 1000 for each replication. 
For each replication we sample two chains starting with 10000 iterations.

We used coverage to assess model fit.
Coverage refers to the frequency with which the computed confidence intervals include the true values of
parameters.
We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for bayesian GLMM)
to evaluate coverage.
We also compare computational speed between different platforms.

@@

The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$). 
Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} but not others.

In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
\mli{refer back to Schielzeth and Forstmeier}

@@

Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
\mli{either keep it or take it out, because  will affect other parameters}
The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.  
\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate? Are these singular models? Would they be diagnosable? Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem? (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
\mli{save it for random slope discussion}

However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
Lastly, it takes $\sim$ 2000 seconds for lme4 to fit 1000 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
\mli{put a lower bound to give up}

@@

We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
We have obtained identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
